\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage[english]{babel}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz}
\usetikzlibrary{matrix}
\usepackage[english]{babel}
\usepackage{mathtools}
\usepackage[a4paper, total={6in, 8in}]{geometry}

\title{MMath project}
\author{Saxon Supple\\Supervisor: Johannes Nordstrom}
\date{January 2024}



\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\begin{document}

\maketitle

\section{Recap of MA40040}

\begin{definition}
A \textbf{topological space} $(X,\mathcal{T})$ is a set $X$ together with a family of subsets $\mathcal{T}\subset\mathcal{P}(X)$ of $X$ such that\\
$\emptyset,X\in\mathcal{T}$,\\
If $\{U_\alpha\}_{\alpha\in I}\subset \mathcal{T}$ then $\bigcup_{\alpha\in I}U_\alpha\in\mathcal{T}$,\\
If $U_1,U_2\in\mathcal{T}$ then $U_1\cap U_2\in\mathcal{T}$.\\
$\mathcal{T}$ is said to be a \textbf{topology} on $X$. The elements $U_\alpha\in\mathcal{T}$ are called the \textbf{open subsets} of X.
\end{definition}

\begin{definition}
Let $(X,\mathcal{T})$ and $(Y,\mathcal{T}')$ be topological spaces. A map $f\colon X\to Y$ is \textbf{continuous} if\\
$f^{-1}(U)\in\mathcal{T}$, $\forall U\in\mathcal{T}'$.
\end{definition}

\begin{definition}
Let $X$ and $Y$ be topological spaces. A continuous map $\phi\colon X\to Y$ is a homeomorphism if it has a continuous inverse $\phi^{-1}\colon Y\to X$.
\end{definition}

\begin{definition}
Let $X$ be a topological space. A \textbf{path} is a continuous map $\gamma$ from $[0,1]$ to $X$. If $\gamma(0)=\gamma(1)$, then $\gamma$ is a \textbf{loop}.
\end{definition}

\begin{definition}
Let $X$ be a topological space and $\gamma_0,\gamma_1\colon[0,1]\to X$ be paths with the same end points. $\gamma_0$ is \textbf{based homotopic to} $\gamma_1$ if there is a continuous map $F\colon[0,1]\times[0,1]\to X$ such that\\
$F(t,0)=\gamma_0(t),t\in[0,1]$\\
$F(t,1)=\gamma_1(t),t\in[0,1]$\\
$F(0,s)=\gamma_0(0),s\in[0,1]$\\
$F(1,s)=\gamma_0(1),s\in[0,1]$\\
We say that $F$ is a \textbf{based homotopy from} $\gamma_0$ \textbf{to} $\gamma_1$ and write $\gamma_0\sim\gamma_1$.
\end{definition}
\begin{lemma}
The based homotopy relation $\sim$ is an equivalence relation.
\end{lemma}

\begin{definition}
Let $\alpha,\beta\colon[0,1]\to X$ be paths in $X$ such that $\alpha(1)=\beta(0)$. The product $\alpha\cdot\beta\colon[0,1]\to X$ is defined as:\\
$(\alpha\cdot\beta)(t)=\alpha(2t)$ for $t\in[0,\frac{1}{2}]$ and $\beta(2t-1)$ otherwise.
\end{definition}

\begin{theorem}
Let $\alpha_0,\alpha_1\colon[0,1]\to X$ be paths from $x_0$ to $y_0$ in $X$ and let $\beta_0,\beta_1\colon[0,1]\to X$ be paths from $y_0$ to $z_0$ in $X$. If $\alpha_0\sim\alpha_1$ and $\beta_0\sim\beta_1$, then $\alpha_0\cdot\beta_0\sim\alpha_1\cdot\beta_1$.
\end{theorem}

This gives a well-defined multiplication of based homotopy classes of loops where if $\alpha$ and $\beta$ are loops based at $x_0$, then we set $[\alpha]\cdot[\beta]=[\alpha\cdot\beta]$.

\begin{definition}
Let $X$ be a topological space and $x_0\in X$. The \textbf{fundamental group} is $\pi_1(X,x_0)=\{[\gamma]:\gamma\text{ is a loop based at }x_0\}$.
\end{definition}
This is a group with identity represented by the constant loop and $[\alpha]$ having inverse $[\bar \alpha]$, where $\bar \alpha(t)=\alpha(1-t)$.

\begin{theorem}
If $X$ is path-connected and $x_0,y_0\in X$, then $\pi_1(X,x_0)\cong\pi_1(X,y_0)$.
\end{theorem}
We can then refer to the fundamental group merely as $\pi_1(X)$.

\begin{definition}
Two continuous maps $\phi_0,\phi_1\colon X\to Y$ of topological spaces are \textbf{homotopic} if there exists a continuous map $F\colon X\times I\to Y$ such that\\
$F(x,0)=\phi_0(X)$, $F(x,1)=\phi_1(x)$, $\forall x\in X$.\\
In this case, we write: $\phi_0\simeq\phi_1$ and say that $F$ is a \textbf{homotopy from} $\phi_0$ \textbf{to} $\phi_1$.
\end{definition}



\begin{definition}
Let $X$ and $Y$ be topological spaces. If there exist continuous maps $\phi\colon X\to Y$ and $\psi\colon Y\to X$ such that $\phi\circ\psi\simeq\text{id}_Y$ and $\psi\circ\phi\simeq\text{id}_X$, then $X$ and $Y$ are said to be \textbf{homotopy-equivalent}. In this case, we write $X\simeq Y$.
\end{definition}

\begin{theorem}
Homotopy-equivalent path-connected topological spaces have isomorphic fundamental groups.
\end{theorem}

\begin{definition}
Let $G\times E\to E,(g,e)\mapsto g\cdot e$ be a left action of a group $G$ on a topological space $E$.\\
We say that the action is \textbf{continuous} if the left multiplication map $L_g\colon E\to E,e\mapsto g\cdot e$ is continuous for all $g\in G$.\\
We say that the action is a \textbf{covering space action} if it is continuous and every $e\in E$ has an open neighbourhood $e\in S\subset E$ such that $gS\cap S=\emptyset$ for all non-trivial elements $g\in G$.
\end{definition}

\begin{theorem}
Let $G$ be a group acting via a covering space action on a simply connected topological space $E$. Then $\pi_1(E/G,[e])\cong G$ for any $[e]\in E/G$.
\end{theorem}

\section{Singular homology}
\begin{definition}
A \textbf{k-simplex} is the convex hull\\
$[v_0,...,v_k]:=\{\sum_{i=0}^k\lambda_iv_i:\lambda_i\in [0,1],\sum_{i=0}^k\lambda_i=1\}$\\
of $k+1$ linearly independent vectors $v_0,...,v_k\in\mathbb{R}^n$.\\
The \textbf{standard k-simplex} is\\$\Delta^k=[e_0,...,e_k]=\{(t_0,...,t_k)\in\mathbb{R}^k:t_i\geq 0, \sum_{i=0}^kt_i=1\}$\\
The vectors $v_0,...,v_k$ define a \textbf{framing} $F_{v_0,...,v_k}\colon\Delta^k\rightarrow [v_0,...,v_k],(t_i)\mapsto \sum_{i=0}^kt_iv_i$.
\end{definition}


\noindent For example $\Delta^0$ is a point, $\Delta^1$ a line segment, $\Delta^2$ a triangle and $\Delta^3$ a tetrahedron.

\begin{definition}
The \textbf{faces} of $[v_0,...,v_k]$ are the (k-1)-simplices $[v_0,...,\overset{\wedge}{v_i},...,v_k]$ where $\overset{\wedge}{v_i}$ means omit the entry $v_i$.
\end{definition}

\begin{definition}
Let $X$ be a topological space. A \textbf{singular k-simplex} in X is a continuous map $\sigma\colon\Delta^k\to X$. The \textbf{k-th chain group} of $X$ is the abelian group\\
$C_k(X):=\{\sum_{i=0}^kn_i\sigma_i:k\in\mathbb{N}_0,n_i\in\mathbb{Z},\sigma_i \text{ is a k-simplex}\}$ where the elements are called \textbf{k-chains}.
\end{definition}
\begin{remark}
$C_k(X)$ is an example of a \textbf{free abelian group}; that is, an abelian group whose elements are of the form $\sum_in_is_i$ where each $s_i$ belongs to some set $S$.
\end{remark}


\begin{definition}
The \textbf{boundary map} $\partial=\partial_k\colon C_k(X)\to C_{k-1}(X)$ is given by\\
$\partial\sigma=\sum_{i=0}^k(-1)^i\sigma\circ F_{e_0,...,\overset{\wedge}{e_i},...,e_k}\in C_{k-1}(X)$\\
on single simplices and extended linearly to all of $C_k(X)$.\\
A k-chain $c\in C_k(X)$ is a \textbf{k-cycle} if $\partial(c)=0$.
\end{definition}
\noindent Intuitively this definition means that the boundary map decomposes $\sigma$ into a sum of maps to its faces--or rather the images of faces of $\Delta^k$--taking into account orientation.


\begin{lemma}
$\partial\circ\partial\colon C_k(X)\to C_{k-2}(x)$ is zero.
\end{lemma}
\begin{proof}
Let $\sigma\in C_k(X)$.\\
$\partial\circ\partial (\sigma)=\partial(\sum_{i=0}^k(-1)^i\sigma\circ F_{e_0,...,\overset{\wedge}{e_i},...,e_k})$\\
$=\sum_{i=0}^k(-1)^i\partial\circ\sigma\circ F_{e_0,...,\overset{\wedge}{e_i},...,e_k}$ by linearity\\
$=\sum_{i=0}^k(\sum_{j<i}(-1)^{i+j}\sigma\circ F_{e_0,...,\overset{\wedge}{e_j},...,\overset{\wedge}{e_i},...,e_k} + \sum_{i<j}(-1)^{i+j-1}\sigma\circ F_{e_0,...,\overset{\wedge}{e_i},...,\overset{\wedge}{e_j},...,e_k})$ (since $e_j$ gets shifted forward one spot when $e_i$ is removed first)\\
$=\sum_{i=0}^k(\sum_{j<i}(-1)^{i+j}\sigma\circ F_{e_0,...,\overset{\wedge}{e_j},...,\overset{\wedge}{e_i},...,e_k} + \sum_{j<i}(-1)^{i+j-1}\sigma\circ F_{e_0,...,\overset{\wedge}{e_j},...,\overset{\wedge}{e_i},...,e_k})$
(interchanging $j$ with $i$ in the second term)\\
$=\sum_{i=0}^k(\sum_{j<i}((-1)^{i+j}+(-1)^{1+j-1})\sigma\circ F_{e_0,...,\overset{\wedge}{e_j},...,\overset{\wedge}{e_i},...,e_k})=0$
\end{proof}

\noindent A simple example of this fact is that the boundary of a ball is a sphere, which in turn has no boundary.\\
It also implies that $\text{im}(\partial:C_{k+1}(X)\rightarrow C_k(X))\in\text{ker}(\partial:C_{k}(X)\rightarrow C_{k-1}(X))$,
allowing the following to be well-defined.
\begin{definition}
The \textbf{singular homology} of $X$ is the collection of abelian groups
$H_k(X)=\frac{\text{ker}(\partial:C_{k}(x)\rightarrow C_{k-1}(X))}{\text{im}(\partial:C_{k+1}(x)\rightarrow C_k(X))}$
\end{definition}

\noindent The utility of the homology groups are that homotopy-equivalent spaces have isomorphic homology groups, thereby giving a way to prove that two topological spaces aren't homotopy-equivalent. To prove this, we must first construct a functor sending a continuous map between topological spaces to a homomorphism between their homology groups.\\

\noindent To give a more general result we shall allow the chain groups to have coefficients in any abelian group $G$ as opposed to merely $\mathbb{Z}$.
\begin{definition}
The k-th singular chain group with coefficients in an abelian group $G$ is $C_k(X;G):=\{\sum_{i=1}^kn_i\sigma_i:k\in\mathbb{N}_0,n_i\in G,\sigma_i \text{ is a k-simplex}\}$ The boundary map $\partial\colon C_k(X;G)\to C_{k-1}(X;G)$ and k-th homology group with coefficients in $G$, $H_k(X;G)$, are then defined as above while replacing $\mathbb{Z}$ with $G$.
\end{definition}

\noindent Let $f\colon X\to Y$ be a continuous map between topological spaces. Given a k-simplex $\sigma\colon\Delta^k\to X$ in $X$, $f\circ\sigma$ is a k-simplex in $Y$. This gives rise to a homomorphism $f_\#\colon C_k(X;G)\to C_k(Y;G):\sum_{i=0}^kn_i\sigma_i\mapsto\sum_{i=0}^kn_if\circ\sigma_i$.
\begin{definition}
This induces a \textbf{push-forward homomorphism} $f_*\colon H_k(X;G)\to H_k(Y;G)$ sending $[c]$ to $[f_\#(c)]$.
\end{definition}
\noindent To show that this is well-defined, note that $\partial f_\#=f_\#\partial$ (in other words $f_\#$ is a chain map) and so $f_\#$ maps closed chains to closed chains. Furthermore suppose $[a-b]=[0]$. Then there exists a $\sigma\in C_{k+1}(X;G)$ such that $a-b=\partial\sigma\implies f_\#(a-b)=f_\#\partial\sigma=\partial f_\#\sigma\implies [f_\#(a)-f_\#(b)]=[0]=f_*([a])-f_*([b])$ as required.

\noindent This argument applies more generally to show that all chain maps induce a well-defined map on homology groups.

\begin{proposition}
The push-forward is a functor.
\end{proposition}
\begin{proof}
Let $X$, $Y$ and $Z$ be topological spaces.
Clearly $\text{id}_{X*}=\text{id}_{H_k(X;G)}$.\\
Now let $f\colon Y\to Z$ and $g\colon X\to X$ be continuous maps. Given $[c]=[\sum_{i=0}^kn_i\sigma_i]\in H_k(X;G)$, $(f\circ g)_*([c])=[\sum_{i=0}^kn_if\circ g(\sigma_i)]=f_*([\sum_{i=0}^kn_ig(\sigma_i)])=f_*\circ g_*([c])$ so $(f\circ g)_*=f_*\circ g_*$.
\end{proof}

\noindent This then implies the standard results from category theory such as the push-forward sending commutative diagrams to commutative diagrams and homeomorphisms to group isomorphisms.

\begin{lemma}
If $f,g\colon X\to Y$ are homotopic continuous maps between topological spaces then $f_*=g_*\colon H_k(X)\to H_k(Y)$
\end{lemma}
\begin{proof}
Let $F\colon X\times I\to Y$ be a homotopy from $f$ to $g$. Given a singular n-simplex $\sigma\colon\Delta^n\to X$ we can define $\sigma\times\text{id}\colon\Delta^n\times I\to X\times I$ and $F_\#(\sigma\times\text{id})\colon\Delta^n\times I\to Y$. We now decompose $\Delta^n\times I$ as the union of $n+1$-simplices. To do this write $\Delta^n\times I$ as $[a_0,...,a_n,b_0,...,b_n]$ where $a_i=(e_i,0)$ and $b_i=(e_i,1)$. Then $\Delta^n\times I$ is the union of the $n+1$-simplices $\alpha_i=[a_0,...,a_i,b_i,...,b_n]$. Now define the "prism map" $P\colon C_n(X)\to C_{n+1}(Y)$ by $P(\sigma)=\sum_i(-1)^iF\circ(\sigma\times\text{id})|_{\alpha_i}$. $\partial P(\sigma)=\sum_i(-1)^i[\sum_{j\leq i}(-1)^jF\circ(\sigma\times\text{id})|_{[a_0,...,\overset{\wedge}{a_j},...,a_i,b_i,...,b_n]}+\sum_{j> i}(-1)^{(j+1)}F\circ(\sigma\times\text{id})|_{[a_0,...,a_i,b_i,...,\overset{\wedge}{b_j},...,b_n]}]$. The sum of terms with $i=j$ is $F\circ(\sigma\times\text{id})|_{[\overset{\wedge}{a_0},b_0,...,b_n]}-F\circ(\sigma\times\text{id})|_{[a_0,\overset{\wedge}{b_0},b_1,...,b_n]}+F\circ(\sigma\times\text{id})|_{[a_0,\overset{\wedge}{a_1},b_1,...,b_n]}-F\circ(\sigma\times\text{id})|_{[a_0,a_1,\overset{\wedge}{b_1},...,b_n]}+...+F\circ(\sigma\times\text{id})|_{[a_0,...,\overset{\wedge}{a_n},b_n]}-F\circ(\sigma\times\text{id})|_{[a_0,...,a_n,\overset{\wedge}{b_n}]}=F\circ(\sigma\times\text{id})|_{[\overset{\wedge}{a_0},b_0,...,b_n]}-F\circ(\sigma\times\text{id})|_{[a_0,...,a_n,\overset{\wedge}{b_n}]}=F\circ(\sigma\times\text{id})|_{[b_0,...,b_n]}-F\circ(\sigma\times\text{id})|_{[a_0,...,a_n]}$ as a telescoping sum which is then $=g\circ\sigma-f\circ\sigma=(g_\#-f_\#)(\sigma)$. The terms with $i\neq j$ can be identified with $-P(\partial\sigma)$ so that $\partial P\sigma=-P(\partial\sigma)+g_\#(\sigma)-f_\#(\sigma)\implies\partial P + P\partial=g_\#-f_\#$. $\partial(\partial P + P\partial)=\partial P\partial=(\partial P + P\partial)\partial$ so $(\partial P + P\partial)$ is a chain map inducing a homomorphism on homology groups. Furthermore given a cycle $\sigma$ we have $(\partial P + P\partial)(\sigma)=\partial P(\sigma) + P\partial(\sigma)=\partial P(\sigma)$ so $\partial P(\sigma)$ sends cycles to boundaries thereby inducing the zero map on homology groups. Thus $g_*-f_*=0$ or $f_*=g_*$.
\end{proof}


\noindent We are now ready to prove that homology is a topological invariant.
\begin{proposition}
If $X$ and $Y$ are homotopy-equivalent then $H_k(X;G)\cong H_k(Y;G)$
\end{proposition}
\begin{proof}
We have continuous maps $f\colon X\to Y$ and $g\colon Y\to X$ such that $g\circ f\simeq\text{id}_X$ and $f\circ g\simeq\text{id}_Y$ giving $g_*\circ f_*=\text{id}_{H_k(X;G)}$ and $f_*\circ g_*=\text{id}_{H_k(Y;G)}$. Thus $f_*$ has a left and right inverse so is a bijection so a group isomorphism.
\end{proof}

\begin{lemma}
If $X$ is path-connected then $H_0(X)=\mathbb{Z}$.
\end{lemma}
\begin{proof}
A singular 0-simplex is a map from a point to a point $x\in X$. let $c=\sum n_i\sigma_i$ be a 0-chain. $\partial c=0$ so $\text{ker}(\partial:C_{0}(x)\rightarrow C_{-1}(X))=C_0(X)$ giving $H_0(X)=\frac{C_0(X)}{\partial C_1(X)}$. Let $x,y\in X$ and let $\sigma_x$ and $\sigma_y$ be their corresponding simplicial 0-simplices. $X$ is path-connected so there exists a simplicial 1-simplex $\sigma$ with $\partial\sigma=\sigma_x-\sigma_y$. Define $f\colon C_0(X)\to\mathbb{Z}:\sum n_ix_i\mapsto\sum n_i$. Clearly $\partial C_1(X)\in \text{ker}f$. Let $z=\sum_{i=0}^ka_i\sigma_i\in\text{ker}f$. Express $a_i\sigma_i$ as $\sum_{j=1}^{|a_i|}a_{ij}\sigma_i$ where $a_{ij}$ is either all $1$ or all $-1$. Thus $z=\sum_{j=1}^rb_j\tau_j$ where $\tau_j\in\{\sigma_0,...,\sigma_k\}$ and $b_j=\pm1$. $|\{j:b_j=1\}|=|\{j:b_j=-1\}|$ so $r$ is even and we can express $z$ as a sum of terms of the form $(\sigma_\alpha - \sigma_\beta)\in\partial C_1(X)$, giving $\text{ker}f\in\partial C_1(X)$. Thus $\text{ker}f\in\partial C_1(X)$. $f$ is surjective so the result follows by the first isomorphism theorem.
\end{proof}

\begin{lemma}
$H_k(\{\text{pt}\})=\begin{cases}
       \mathbb{Z} &\quad\text{if }k=0 \\
       0 &\quad\text{otherwise} \\ 
     \end{cases}$
\end{lemma}
\begin{proof}
There is only one k-simplex $\sigma_k\colon\Delta^k\rightarrow \{\text{pt}\}$, the constant map. Thus $C_k(\{\text{pt}\})=\{n\sigma_k:n\in\mathbb{Z}\}=\mathbb{Z}$.
Also $\partial_k(n\sigma_k)=\sum_{i=0}^k(-1)^i n\sigma_{k-1}$ which is zero if $k$ is odd and $n\sigma_{k-1}$ if $k$ is even. Thus the induced map on the coefficients of the boundary is $0$ if $k$ is odd and id if $k$ is even. The Homology groups will then be $\frac{\mathbb{Z}}{\mathbb{Z}}=0$ or $\frac{\{0\}}{\{0\}}=0$, except for $H_0$ which is $\mathbb{Z}$ since $\{\text{pt}\}$ is path-connected.
\end{proof}


\begin{definition}
The \textbf{commutator subgroup} of a group $G$, denoted $[G,G]$, is the subgroup generated by all elements, called $commutators$, of the form $ab(ba)^{-1}$ for any $a$ and $b$ in $G$.
\end{definition}

\begin{lemma}
$[G,G]$ is a normal subgroup.
\end{lemma}
\begin{proof}
Let $g\in G,aba^{-1}b^{-1}\in [G,G]$.\\
$g^{-1}(aba^{-1}b^{-1})g=g^{-1}agg^{-1}bgg^{-1}a^{-1}gg^{-1}b^{-1}g$\\
$=(g^{-1}ag)(g^{-1}bg)(g^{-1}a^{-1}g)(g^{-1}b^{-1}g)$\\
$=(g^{-1}ag)(g^{-1}bg)(g^{-1}ag)^{-1}(g^{-1}bg)^{-1}\in[G,G]$.
\end{proof}
\begin{definition}
The \textbf{abelianization} $G^{\text{ab}}$ of a group $G$ is $G/[G,G]$.
\end{definition}
\noindent As the name suggests, $G^{\text{ab}}$ is abelian since we quotiented out by the relation that $ab=ba$ for all $a,b\in G$. If $G$ is already abelian then $[G,G]$ is trivial so $G^{\text{ab}}=G$.

\begin{lemma}
Given groups $G$ and $H$, where $H$ is abelian, and a group homomorphism $h\colon G\to H$, the induced homomorphism $h'\colon G^{\text{ab}}\to H:[x]\mapsto h(x)$ is well-defined.
\end{lemma}
\begin{proof}
Let $[x]=[y]\in G^{\text{ab}}$ so that $x=y\prod_ia_ib_ia_i^{-1}b_i^{-1}$ for $a_i,b_i\in G$.
Then $h'([x])=h(y\prod_ia_ib_ia_i^{-1}b_i^{-1})=h(y)\prod_ih(a_i)h(b_i)h(a_i)^{-1}h(b_i)^{-1}=h(y)=h'([y])$ since $H$ is abelian.
\end{proof}

\begin{lemma}
Let $G$ be a group. If $w$ is a word in $G$ such that for any $g\in G$ the exponents of $g$ in $w$ sum to zero then $w\in[G,G]$ 
\end{lemma}
\begin{proof}
Consider the quotient map $\pi\colon G\to G/[G,G]:g\mapsto[g]$.
$G/[G,G]$ is abelian so the letters in $[w]$ can be reordered such that $[w]=\sum n_i[a_i]=0$ implying $w\in\text{ker }\pi=[G,G]$. 
\end{proof}

\begin{theorem}
(Hurewicz). If $X$ is path-connected then $H_1(X)\cong\pi_1(X)^{\text{ab}}$.
\end{theorem}
\begin{proof}
First note that an element of $\pi_1(X,x_0)$ is of the form $[\gamma]$ for $\gamma\colon [0,1]\to X$ a loop based at $x_0$. $[0,1]=\Delta^1$ so $\gamma$ is a 1-simplex and $\partial(\gamma)=x_0-x_0$ making $\gamma$ closed.
We can then define $h\colon\pi_1(X,x_0)\to H_1(X)$ sending an element $[\gamma]\in\pi_1(X,x_0)$ to $[c_\gamma]\in H_1(X)$ where $c_\gamma$ is $\gamma$ as an element of $C_1(X)$. To show that $h$ is well-defined let $[a]=[b]\in\pi_1(X,x_0)$ and let $F\colon [0,1]^2\to X$ be a based-homotopy from $a$ to $b$. Dividing $[0,1]^2$ into two triangles and parametrising each by $\Delta^2$ allows F to be expressed as $c_F\in C_2(X)$ with $\partial c_F=c_a-c_b\implies [c_a]=[c_b]$ as required.

\noindent \textbf{Next we show that $h$ is a homomorphism.}\\
Let $[f],[g]\in\pi_1(X,x_0)$ and Define a singular 2-simplex $\sigma\colon\Delta^2\to X$ where $\sigma_{|[e_0,e_1]}=f$, $\sigma_{|[e_1,e_2]}=g$ and $\sigma_{|[e_0,e_2]}=f\cdot g$. Then $\partial(\sigma)=g-f\cdot g+f$ so $h([f][g])=h([f])+h([g])$ making $h$ a homomorphism.

\noindent \textbf{We now show that $h$ is surjective.}\\
For the sake of lucidity we will write $[\gamma]$ for $h([\gamma])$ instead of $[c_\gamma]$\\
Let $[\lambda]\in H_1(X)$ where $\lambda=\sum_{i=0}^k n_i\sigma_i$.
$\partial (\lambda)=0=\sum_{i=0}^k n_i(\sigma_i(1)-\sigma_i(0))$.\\
Let $\Lambda=\{\sigma_i(0)\}\cup\{\sigma_i(1)\}$ be the collection of end-points of the $\sigma_i$s.
For each $p\in\Lambda$ pick a path $\beta_p\colon [0,1]\to X$ from $x_0$ to $p$ (which is possible because $X$ is path-connected). Then $\sum_{i=0}^k n_i(\sigma_i(1)-\sigma_i(0))=\sum m_pp=0$ where $m_p$ is the sum of the coefficients of $p$ in the sum. This implies that $m_p=0 \forall p$. Now consider the loop $\eta_i=\beta_{\sigma_i(0)}\cdot\sigma_i\cdot\beta_{\sigma_i(1)}^{-1}$ based at $x_0$. Then $h([\eta_0^{n_0}\cdot\eta_1^{n_1}...\cdot\eta_k^{n_k}])=[n_0\eta_0+...+n_k\eta_k]=[\sum_{i=0}^kn_i(\beta_{\sigma_i(0)}+\sigma_i-\beta_{\sigma_i(1)})]=[\sum_{i=0}^kn_i\sigma_i+\sum_{i=0}^kn_i(\beta_{\sigma_i(0)}-\beta_{\sigma_i(1)})]=[\sum_{i=0}^kn_i\sigma_i-\sum m_p\beta_p]=[\sum_{i=0}^kn_i\sigma_i]=[\lambda]$ since $m_p=0$.\\
\noindent\textbf{We now find the kernel}\\
Let $[\gamma]\in\pi_1(X,x_0)$ such that $h([\gamma])=[0]$.
Then $\gamma$ is exact so there exists a finite sum of singular 2-simplices $\sum_{i=0}^kn_i\sigma_i$ such that $\gamma=\partial(\sum_{i=0}^kn_i\sigma_i)=\sum_{i=0}^kn_i\partial(\sigma_i)=\sum_{i=0}^kn_i(\lambda_i+\mu_i+\nu_i)$. Let $\Lambda=\{\lambda_i\}\cup\{\mu_i\}\cup\{\nu_i\}$ and for each end-point $p$ of a singular 1-simplex in $\Lambda$ let $\beta_p$ be a path from $x_0$ to $p$. For each $\theta\in\Lambda$ let $m_\theta$ be the sum of coefficients of $\theta$ in $\sum_{i=0}^kn_i(\lambda_i+\mu_i+\nu_i)$. This sum is equal to $\gamma$ thus $m_\theta$ is $1$ if $\theta=\gamma$ and $0$ otherwise. For each singular 2-simplex $\sigma_i$ the loops $\beta_{\lambda_i(0)}\cdot\lambda_i\cdot\beta_{\lambda_i(1)}^{-1}$, $\beta_{\mu_i(0)}\cdot\mu_i\cdot\beta_{\mu_i(1)}^{-1}$ and $\beta_{\nu_i(0)}\cdot\nu_i\cdot\beta_{\nu_i(1)}^{-1}$ have null-homotopic concatenations. Let $\eta_i=\beta_{\partial\sigma_i(0)}\cdot\partial\sigma_i\cdot\beta_{\partial\sigma_i(1)}^{-1}$. Then $h([\eta_0^{n_0}\cdot...\cdot\eta_k^{n_k}\cdot\gamma^{-1}])=h([\gamma])h([\gamma^{-1}])=[\gamma^{-1}]$. Now write each $[\eta_i]$ as $[\beta_{\lambda_i(0)}\cdot\lambda_i\cdot\beta_{\lambda_i(1)}^{-1}][\beta_{\mu_i(0)}\cdot\mu_i\cdot\beta_{\mu_i(1)}^{-1}][\beta_{\nu_i(0)}\cdot\nu_i\cdot\beta_{\nu_i(1)}^{-1}]$. Then for each $\theta\in\Lambda$ we have that the sum of exponents of $[\beta_{\theta(0)}\cdot\theta\cdot\beta_{\theta(1)}^{-1}]$ is $m_\theta$ The exponents of each $[\beta_{\theta(0)}\cdot\theta\cdot\beta_{\theta(1)}^{-1}]$ in $[\eta_0^{n_0}\cdot...\cdot\eta_k^{n_k}\cdot\gamma^{-1}]$ is then 0 so by the above lemma $[\gamma^{-1}]\in[\pi_1(X,x_0),\pi_1(X,x_0)]$ meaning $[\gamma]$ is as well. Thus $\text{ker }h\subseteq [\pi_1(X,x_0),\pi_1(X,x_0)]$. We also have that $[\pi_1(X,x_0),\pi_1(X,x_0)]\subseteq \text{ker }h$ since $H_1(X)$ is abelian so $\text{ker }h=[\pi_1(X,x_0),\pi_1(X,x_0)]$. The result then follows by the first isomorphism theorem.
\end{proof}

\begin{corollary}
$H_1(S^n)=$$\begin{cases}
       \mathbb{Z} &\quad\text{if }n=1 \\
       0 &\quad\text{otherwise} \\ 
     \end{cases}$\\
$H_1(T^2)=\mathbb{Z}^2$\\
$H_1(\mathbb{R}P^n)=$$\begin{cases}
       \mathbb{Z} &\quad\text{if }n=1 \\
       \mathbb{Z}_2 &\quad\text{otherwise} \\ 
     \end{cases}$\\
$H_1(K)=\mathbb{Z}\bigoplus\mathbb{Z}_2$
\end{corollary}

\section{Cohomology}

\begin{definition}
Let $R$ be a commutative ring with a unit. An \textbf{R-module} $M$ is an abelian group with a map $f\colon R\times M\to M$ such that if $f(r,m)$ is denoted by $rm$ then $(r_1r_2)m=r_1(r_2m)$ and $1m=m$. We usually abbreviate R-modules to modules.
\end{definition}
\begin{remark}
More specifically we defined a \textbf{left R-module}. If instead multiplication were defined as $r_1(r_2m)=(r_2r_1)m$ then $M$ would be a \textbf{right R-module}.
\end{remark}

\begin{definition}
Let $M$ and $N$ be R-modules. An \textbf{R-module homomorphism} $f\colon M\to N$ is a homomorphism of abelian groups such that $f(rm)=rf(m)$ 
\end{definition}

\begin{example}
An abelian group $G$ is a $\mathbb{Z}$-module given by $ng=$$\begin{cases}
       \sum_{i=1}^ng &\quad\text{if }n>0 \\
       0 &\quad\text{if }n=0 \\
       -\sum_{i=1}^{-n}g &\quad\text{if }n<0 \\
     \end{cases}$\\
\end{example}

\begin{definition}
Let $A$ and $B$ be R-modules. Their \textbf{direct sum} $A\bigoplus B$ is the module $A\times B$ under component-wise operations i.e. $(a_1,b_1)+(a_2,b_2)=(a_1+a_2,b_1+b_2)$ and $r(a,b)=(ra,rb)$. This can be extended to any finite collection of R-modules.
\end{definition}

\begin{definition}
A bilinear map $h\colon M\times N\to P$ is a map such that $h(m_1+m_2,n)=h(m_1,n)+h(m_2,n)$, $h(m,n_1+n_2)=h(m,n_1)+h(m,n_2)$ and $h(rm,n)=h(m,rn)$ for $r\in R$.
\end{definition}

\begin{definition}
Given a ring $R$ and set $S$, we define the module $R^S$ to be functions from $S$ to $R$, with addition and multiplication defined piecewise.
\end{definition}


\begin{definition}
Given a module M, we say that $H\subseteq M$ is a \textbf{submodule} if $rh\in H\forall r\in R,h\in H$ and $H$ is itself a module after inheriting the addition and multiplication maps of $M$.
\end{definition}

\begin{definition}
Given a module $M$ and submodule $N$, the \textbf{quotient R-module of M by N} is the quotient group $M/N$ with multiplication given by $r(a+N)=ra+N\forall r\in R,a+N\in M/N$.
\end{definition}

\begin{definition}
Let $M$ and $N$ be modules. Let $I$ denote the submodule of $M\times N$ generated by the set of elements of the form $(\lambda m_1+m_2,n)-\lambda(m_1,n)-(m_2,n)$, $(m,\lambda n_1+n_2)-\lambda(m,n_1)-(m,n_2)$ for $m,m_1,m_2\in M$, $n,n_1,n_2\in N$ and $\lambda\in R$. Then the \textbf{tensor product} of $M$ and $N$ is $M\bigotimes_R N=R^{M\times N}/I$ whose elements are represented as $m\otimes n$ for $m\in M$, $n\in N$.
\end{definition}

\noindent This gives rise to the relations\\ $(m_1+m_2)\otimes n=m_1\otimes n+m_2\otimes n$,\\
$m\otimes(n_1+n_2)=m\otimes n_1+m\otimes n_2$ and\\
$(rm)\otimes n=m\otimes (rn)=r(m\otimes n)$ for $r\in R$.

\begin{example}
If $p$ and $q$ are coprime then $\mathbb{Z}_p\bigotimes_\mathbb{Z}\mathbb{Z}_q=0$ since there exist $m,n\in\mathbb{Z}$ such that $mp+nq=1$ (Bezout's lemma) so $a\otimes b=(mp+nq)a\otimes b=(mpa)\otimes b + a\otimes(nqb)=0\forall a\otimes b\in\mathbb{Z}_p\bigotimes_\mathbb{Z}\mathbb{Z}_q$.
\end{example}
\begin{example}
$C_k(X;G)=C_k(X)\bigotimes_\mathbb{Z}G$.
\end{example}

\begin{example}
$\mathbb{Z}_m\bigotimes\mathbb{Z}_n=\mathbb{Z}_{gcd(m,n)}$.
\end{example}
\begin{proof}
We have the first isomorphism for modules that given a module homomorphism $f:M\to N$, $M/\text{ker }f\cong\text{im }f$ (the proof is analogous to the cases for groups, rings and vector spaces).
So, define a module homomorphism $g\colon\mathbb{Z}\to\mathbb{Z}_m\bigotimes\mathbb{Z}_n:z\mapsto z(1\otimes 1)$. $g$ is surjective so $\text{im }g=\mathbb{Z}_m\bigotimes\mathbb{Z}_n$. Clearly $<\text{gcd}(m,n)>\subseteq\text{ker }g$. Now let $x\notin <\text{gcd}(m,n)>$. Then $x=s\text{gcd}(m,n)+k$ with $0<k<\text{gcd}(m,n)$ by division with remainder. $g(x)=s\text{gcd}(m,n)(1\otimes 1)+g(k)=k(1\otimes 1)\neq 0\implies x\notin\text{ker }g$ so $x\in\text{ker }g\implies x\in<\text{gcd}(m,n)>$ giving $<\text{gcd}(m,n)>=\text{ker }g$.
\end{proof}


\begin{definition}
Let M and N be modules over a ring R. The set of all module homomorphisms from M to N is denoted by $\text{Hom}_R(M,N)$. It is an abelian group under pointwise addition and a module if R is commutative.
\end{definition}

\begin{definition}
Let $G$ be an abelian group. The \textbf{torsion subgroup} $T(G)$ is the subgroup consisting of all elements of $G$ with finite order. For a prime number p, the \textbf{p-torsion} subgroup $T_p(G)$ is the subgroup consisting of all elements with order a power of p. The \textbf{free part} of $G$, denoted $F(G)$, is the subgroup comprising all elements of infinite order, as well as the identity.
\end{definition}

\noindent We then have that $G=T(G)\bigoplus F(G)$ for any abelian group $G$.


\begin{definition}
Let $X$ be a topological space and let $G$ be an abelian group. The \textbf{kth cochain group with coefficients in G} is $C^k(X;G)=Hom(C_k(X),G)$. 
\end{definition}

\begin{definition}
We define $d\colon C^k(X;G)\to C^{k+1}(X;G)$ to be the dual of $\partial$ in the sense that if $\phi\in C^k(X;G)$ and $c\in C_{k+1}(X)$ then $(d\phi)(c)=\phi(\partial c)\in G$
\end{definition}

\noindent In particular, $(d\circ d\phi)(c)=(d\phi)(\partial c)=\phi(\partial\circ\partial c)=\phi(0)=0$ so $d^2=0$.

\begin{definition}
$c\in C_k(X)$ is a cycle if $\partial c=0 and a boundary if c\in\text{im }\partial$.
$\phi\in C^k(X;G)$ is a cocycle if $d\phi=0$ and a coboundary if $\phi\in\text{im }d$.
\end{definition}

\begin{definition}
The \textbf{kth cohomology group with coefficients in $G$} is $H^k(X;G)=\frac{\text{ker}(d\colon C^k(X;G)\to C^{k+1}(X;G))}{\text{im}(d\colon C^{k-1}(X;G)\to C^k(X;G))}$. When G is $\mathbb{Z}$ we abbreviate it to $H^k(X)$ and call it simply the \textbf{kth cohomology group}.
\end{definition}

\begin{definition}
Let $f\colon X\to Y$ be a continuous map between topological spaces and $G$ be an abelian group. We have a homomorphism $f^\#\colon C^k(Y;G)\to C^k(X;G):A\mapsto f^\# A$ given by $f^\# A(c)=A(f_\#(c))$. This in turn induces the \textbf{pullback homomorphism} $f^*\colon H^k(Y;G)\to H^k(X;G):[A]\mapsto [f^\# A]$.
\end{definition}
\noindent $df^\# A(c)=f^\# A(\partial c)=f^\# dA(c)$ so $df^\#=f^\# d$. Thus $f^\#$ is a chain map so $f^*$ is well-defined.

\begin{lemma}
There is a well-defined mapping $H^k(X;G)\times H_k(X)\to G$ given by $([\phi],[c])=\phi(c)$.
\end{lemma}
\begin{proof}
Let $[\phi]=[\psi]$ so that $\phi-\psi=d\rho$ for some $\rho\in C^{k-1}(X;G)$. Then $(\phi-\psi)(c)=d\rho(c)=\rho(\partial c)=0$ since $c\in\text{ker }\partial$. Thus $\phi(c)=\psi(c)$.
Now let $[a]=[b]$ so that $a=b+\partial e$ for some $e\in C_{k+1}(X)$. $\phi(a)=\phi(b+\partial e)=\phi(b)+\phi(\partial e)=\phi(b)+d\phi(e)=\phi(b)$ since $\phi\in\text{ker }d$.
\end{proof}

\noindent This then give that $f^*$ is dual to $f_*$ in the sense that $f^*[A]([c])=[A](f_*([c]))$ for $[A]\in H^k(Y;G)$ and $[c]\in H_k(X)$.

\begin{proposition}
Pullback is a contravariant functor.
\end{proposition}
\begin{proof}
Let $X$, $Y$ and $Z$ be topological spaces and let $G$ be an abelian group. 
Clearly $\text{id}_X^*=\text{id}_{H^k(X;G)}$.
Now let $f\colon X\to Y$ and $g\colon Y\to Z$ be continuous maps. $(g\circ f)^*([A])=[(g\circ f)^\#(A)]=[f^\#\circ g^\#(A)]=f^*([g^\#(A)])=f^*\circ g^*([A])\forall [A]\in H^k(Z;G)$. Thus $(g\circ f)^*=f^*\circ g^*$.
\end{proof}

\begin{proposition}
Let $X$ and $Y$ be homotopy-equivalent topological spaces. Then $H^k(X;G)\cong H^k(Y;G)$.
\end{proposition}
\begin{proof}
We have continuous maps $f\colon X\to Y$ and $g\colon Y\to X$ such that $g\circ f\simeq\text{id}_X$ and $f\circ g\simeq\text{id}_Y$. $(g\circ f)_*=\text{id}_{H_k(X)}$ so $f^*\circ g^*=(g\circ f)^*=\text{id}_{H^k(X;G)}$ by duality. Similarly $g^*\circ f^*=\text{id}_{H^k(Y;G)}$. Thus $f^*$ is a bijective module homomorphism so an isomorphism.
\end{proof}


\section{Cup product and ring structure of cohomology}
\begin{definition}
Let $R$ be a ring and $X$ be a topological space. The \textbf{cup product} is the product on cochains $C^i(X;R)\times C^j(X;R)\to C^{i+j}(X;R):(\psi,\phi)\mapsto\psi\cup\phi$ where $(\psi\cup\phi)\sigma=\psi(\sigma\circ F_{[e_0,...,e_i]})\phi(\sigma\circ F_{[e_i,...,e_{i+j}]})$ for a singular $i+j$-simplex $\sigma\colon\Delta^{i+j}\to X$ and is extended linearly to $C_{i+j}(X)$.
\end{definition}

\begin{lemma}
$d(\psi\cup\phi)=(d\psi)\cup\phi+(-1)^i\psi\cup d\phi$.
\end{lemma}
\begin{proof}
Let $\sigma\colon\Delta^{i+j+1}\to X$ be a singular simplex.
$d(\psi\cup\phi)(\sigma)=\psi\cup\phi(\partial\sigma)=\psi\cup\phi(\sum_{t=0}^{j+k+1}(-1)^t\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+j+1}]})=\sum_{t=0}^{i+j+1}(-1)^t\psi\cup\phi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+j+1}]})=\sum_{t=0}^{i}(-1)^t\psi\cup\phi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+j+1}]})+\sum_{t=i+1}^{i+j+1}(-1)^t\psi\cup\phi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+j+1}]})=\sum_{t=0}^i(-1)^t\psi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+1}]})\phi(\sigma\circ F_{[e_{i+1},...,e_{i+j+1}]})+\sum_{t=i+1}^{i+j+1}(-1)^t\psi(\sigma\circ F_{[e_0,...,e_{i}]})\phi(\sigma\circ F_{[e_{i},...,\overset{\wedge}{e_t},...,e_{i+j+1}]})=\sum_{t=0}^{i+1}(-1)^t\psi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+1}]})\phi(\sigma\circ F_{[e_{i+1},...,e_{i+j+1}]})+\sum_{t=i}^{i+j+1}(-1)^t\psi(\sigma\circ F_{[e_0,...,e_{i}]})\phi(\sigma\circ F_{[e_{i},...,\overset{\wedge}{e_t},...,e_{i+j+1}]})$.

\noindent We have $(d\psi\cup\phi)(\sigma)=\sum_{t=0}^{i+1}(-1)^t\psi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},...,e_{i+1}]})\phi(\sigma\circ F_{[e_{i+1},...,e_{i+j+1}]})$
and $(-1)^i(\psi\cup d\phi)(\sigma)=\sum_{t=i}^{i+j+1}(-1)^t\psi(\sigma\circ F_{[e_0,...,e_i]})\phi(\sigma\circ F_{[e_i,...,\overset{\wedge}{e_t},...,e_{i+j+1}]})$. When these expressions are added, the last term of the first sum cancels with the first term of the second sum leaving $d(\psi\cup\phi)(\sigma)$.
\end{proof}

\begin{proposition}
There is a well-defined map $H^i(X;R)\times H^j(X;R)\to H^{i+j}(X;R)$ given by $([\psi],[\phi])\mapsto [\psi]\cup [\phi]=[\psi\cup\phi]$.
\end{proposition}
\begin{proof}
$d(\psi\cup\phi)=0\cup\phi+(-1)^i\psi\cup 0=0$ so $\psi\cup\phi \in \text{ker }d$.
Now let $\psi=df$. Then $d(f\cup\phi)=(df)\cup\phi\pm f\cup d\phi=\psi\cup\phi+f\cup0=\psi\cup\phi$. Similarly if $\phi=dg$ then $d((-1)^i\psi\cup g)=\psi\cup\phi$ so adding an element in the image of $d$ keeps $\psi\cup\phi$ in the same equivalence class.
\end{proof}

\begin{definition}
We call the induced map $\cup:H^k(X;R)\times H^l(X;R)\to H^{k+l}(X;R)$ the \textbf{cup product} on cohomology.
\end{definition}

\begin{definition}
A \textbf{graded ring} is a ring such that its additive group is a direct sum of abelian groups.
\end{definition}

\begin{proposition}
$H^*(X;R)=\bigoplus_{n=0}^\infty H^n(X;R)$ is a graded ring with multiplication given by the cup product.
\end{proposition}
\begin{proof}
The direct sum of abelian groups is abelian so $H^*(X;R)$ is abelian as an additive group.
To show that the product is associative and distributive it suffices to show that it is for cochains.\\
For associativity:
$(\psi\cup\phi)\cup\theta(\sigma)=(\psi(\sigma\circ F_{[e_0,...,e_k]})\phi(\sigma\circ F_{e_k,...,e_{k+l}}))\theta(\sigma\circ F_{[e_{k+l},...,e_{k+l+m}]})=\psi(\sigma\circ F_{[e_0,...,e_k]})(\phi(\sigma\circ F_{[e_k,...,e_{k+l}]})\theta(\sigma\circ F_{[e_{k+l},...,e_{k+l+m}]}))=(\psi\cup\phi)\cup\theta(\sigma)$.\\
For distributivity:
$\psi\cup(\phi+\theta)(\sigma)=\psi(\sigma\circ F_{[e_0,...,e_i]})(\phi(\sigma\circ F_{[e_i,...,e_{i+j}]})+\theta(\sigma\circ F_{[e_i,...,e_{i+j}]}))=\psi(\sigma\circ F_{[e_0,...,e_i]})\phi(\sigma\circ F_{[e_i,...,e_{i+j}]})+\psi(\sigma\circ F_{[e_0,...,e_i]})\theta(\sigma\circ F_{[e_i,...,e_{i+j}]})=\psi\cup\phi(\sigma)+\psi\cup\theta(\sigma)$.


\noindent If $R$ has an identity element then there is also an identity element given by $[1]\in H^0(X;R)$ which sends each singular 0-simplex to 1.
\end{proof}

\begin{proposition}
Let $R$ be a commutative ring, $\psi\in H^k(X;R)$ and $\phi\in H^l(X;R)$. Then $\psi\cup\phi=(-1)^{kl}\phi\cup\psi$.
\end{proposition}

\begin{proposition}
Let $f\colon X\to Y$ be a continuous map. Then $f^*(\psi\cup\phi)=f^*(\psi)\cup f^*(\phi)$.
\end{proposition}
\begin{proof}
As before it suffices to check this for cochains. Let $\sigma\colon\Delta^{p+q}\to X$ be a singular simplex. Then $(f^*\psi\cup f^*\phi)(\sigma)=f^*\psi(\sigma\circ F_{[e_0,...,e_p]})f^*\phi(\sigma\circ F_{[e_p,...,e_{p+q}]})=\psi(f\circ\sigma\circ F_{[e_0,...,e_p]})\phi(f\circ\sigma\circ F_{[e_p,...,e_{p+q}]})=(\psi\cup\phi)(f\circ\sigma)=f^*(\psi\cup\phi)(\sigma)$.
\end{proof}

\noindent This makes $f^*$ a ring homomorphism.

\section{Cap product and Poincare duality}

\begin{definition}
Let $X$ be a topological space and $R$ a commutative ring with an identity element. The \textbf{cap product} $C_{i+j}(X;R)\times C^i(X;R)\to C_j(X;R):(c,\phi)\mapsto c\cap\phi$ given by $\sigma\cap\phi=\phi(\sigma\circ F_{[e_0,...,e_i]})\sigma\circ F_{[e_i,...,e_{i+j}]}$ for $\sigma\colon\Delta^j\to X$ a singular simplex and $\phi\in C^{i+j}(X;R)$ and extended linearly.
\end{definition}

\begin{lemma}
$\partial(\sigma\cap\phi)=(-1)^i(\partial\sigma\cap\phi-\sigma\cap d\phi)$.
\end{lemma}
\begin{proof}
\[\partial\sigma\cap\phi=\sum_{t=0}^i(-1)^t\phi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},..,e_{i+1}]})\sigma\circ F_{[e_{i+1},...,e_{i+j}]}+\sum_{t=i+1}^{i+j}(-1)^t\phi(\sigma\circ F_{[e_0,..,e_{i}]})\sigma\circ F_{[e_{i},...,\overset{\wedge}{e_t},...,e_{i+j}]}\]
\[\sigma\cap d\phi=\sum_{t=0}^{i+1}(-1)^t\phi(\sigma\circ F_{[e_0,...,\overset{\wedge}{e_t},..,e_{i+1}]})\sigma\circ F_{[e_{i+1},...,e_{i+j}]}\]
\[\partial(\sigma\cap\phi)=\sum_{t=i}^{i+j}(-1)^{t-i}\phi(\sigma\circ F_{[e_0,..,e_{i}]})\sigma\circ F_{[e_{i},...,\overset{\wedge}{e_t},...,e_{i+j}]}\]
\end{proof}

\begin{proposition}
There is an induced cap product $H_{i+j}(X;R)\times H^i(X;R)\to H_j(X;R)$ which is $R$-linear in each variable.
\end{proposition}
\begin{proof}
The cap product of a cycle and a cocycle is a cycle. $\partial\sigma=0\implies\partial(\sigma\cap\phi)=\pm(\sigma\cap d\phi)$ so the cap product of a cycle and a coboundary is a boundary. $d\phi=0\implies\partial(\sigma\cap\phi)=\pm(\partial\sigma\cap\phi)$ so the cap product of a boundary and a cocycle is a boundary.
\end{proof}

\begin{proposition}
Let $f\colon X\to Y$ be continuous, $c\in C_{i+j}(X;R)$ and $\phi\in C^i(X;R)$. Then $f_*(c)\cap\phi=f_*(c\cap f^*(\phi))$.
\end{proposition}
\begin{proof}
Let $\sigma\colon\Delta^{i+j}\to X$ be a singular simplex. Then $f\sigma\cap\phi=\phi(f\sigma\circ F_{[e_0,...,e_j]})f\sigma\circ F_{[e_j,...,e_{i+j}]}$.
The result follows by linearity.
\end{proof}

\begin{definition}
A topological space $M$ is an \textbf{n-dimensional manifold} if for every $p\in M$ there is an open neighbourhood $U\subseteq M$ containing $p$ and a homeomorphism $f\colon U\to V$ where $V\subseteq\mathbb{R}^n$ is open.
\end{definition}

\begin{definition}
An n-dimensional closed manifold $M$ is \textbf{orientable} if $H_n(M)\cong \mathbb{Z}$.
\end{definition}

\begin{definition}
Let $M$ be an n-dimensional manifold and $R$ a commutative ring with an identity. $M$ is \textbf{R-orientable} if $H_n(M;R)\cong R$. The image of a unit in $R$ is called a \textbf{fundamental class}. An \textbf{orientation} of $M$ is a choice of fundamental class $[M]\in H_n(M)$.
\end{definition}

\begin{proposition}
Let $M$ be an n-dimensional manifold and $R$ a commutative ring with an identity. Then $H_n(M;R)\cong R$ if $M$ is orientable and $T_2R$ otherwise. Furthermore, $H_i(M;R)=0$ for $i>n$.
\end{proposition}

\begin{definition}
Let $M$ and $N$ be closed connected orientable $n$-dimensional manifolds, let $f\colon M\to N$ be continuous and let $[M]$ and $[N]$ be fundamental classes of $X$ and $Y$. Then the \textbf{degree} of $f$ is the image of $f_*([M])$. Or in other words, $f_*([M])=\text{deg}(f)[N]$.
\end{definition}

\begin{proposition}
Let $M$ and $N$ be R-orientable n-dimensional manifolds and $f\colon M\to N$ a homeomorphism. Then $f_*$ sends a fundamental class of $M$ to a fundamental class of $N$.
\end{proposition}
\begin{proof}
Let $[M]\in H_n(M;R)$ be a fundamental class of $M$ and let $\phi\colon H_n(M;R)\to R$ and $\psi\colon H_n(N;R)\to R$ be the module isomorphisms. $\phi([M])$ is a unit so there exists an inverse $\phi([M])^{-1}=\phi([A])$ so that $\psi(f_*([M]))\psi(f_*([A]))=(\psi\circ f_*\circ \phi^{-1})(\phi([M]))(\psi\circ f_*\circ \phi^{-1})(\phi([A]))=(\psi\circ f_*\circ \phi^{-1})(\phi([M])\phi([A]))=(\psi\circ f_*\circ \phi^{-1})(1)=1$. Thus $\psi(f_*([M]))$ is a unit so $f_*([M])$ is a fundamental class.
\end{proof}

\begin{corollary}
The degree of a homeomorphism between closed connected orientable $n$-manifolds is $\pm 1$.
\end{corollary}

\begin{definition}
Let $M$ be an orientable n-dimensional manifold, let $[M]\in H_n(X)$ be a fundamental class and let $f\colon M\to M$ be a homeomorphism. We say that $f$ is \textbf{orientation-preserving} if $f_*([M])=[M]$ and that $f$ is \textbf{orientation-reversing} if $f_*([M])=-[M]$.
\end{definition}


\begin{definition}
A manifold is \textbf{closed} if it is compact and has no boundary.
\end{definition}

\begin{theorem}
(Poincare Duality). If $M$ is a closed $R$-orientable $n$-manifold with fundamental class $[M]\in H_n(M;R)$, then the map $D\colon H^k(M;R)\to H_{n-k}(M;R)$ defined by $D([\alpha])=[M]\cap[\alpha]$ is an isomorphism for all $k$.
\end{theorem}

\begin{proposition}
Let $\psi\in C^l(X;R)$, $\alpha\in C_{k+l}(X;R)$ and $\psi\in C^k(X;R)$. Then $\psi(\alpha\cap\phi)=(\phi\cap\psi)(\alpha)$
\end{proposition}
\begin{proof}
Let $\sigma\colon\Delta^{k+l}\to X$ be a singular simplex. Then $\psi(\sigma\cap\phi)=\psi(\phi(\sigma\circ F_{[e_0,...,e_k]})\sigma\circ F_{[e_k,...,e_{k+l}]})=\phi(\sigma\circ F_{[e_0,...,e_k]})\psi(\sigma\circ F_{[e_k,...,e_{k+l}]})$.
The result then follows by linearity.
\end{proof}

\section{Cell complexes}
\begin{definition}
Let $Y$ be a topological space and let $\phi\colon S^{n-1}\to Y$ be a continuous map. Define a relation on $Y\coprod B^n$ generated by $x\sim y$ if $x\in S^{n-1}\cong\partial B^n$ and $y=\phi(x)$ and let $Y\cup_\phi B^n=Y\coprod B^n/\sim$ with the quotient topology. We call this $Y$ with an n-cell attached via $\phi$.
\noindent As a set, $Y\cup_\phi B^n=Y\coprod e^n$ where the \textbf{n-cell} $e^n$ is the image of the interior of $B^n$.
\end{definition}

\begin{definition}
A \textbf{finite cell complex} $X$ of dimension 0 is a discrete finite set $X^0$ of $0$-cells. A finite cell complex $X$ of dimension n is the result of attaching a finite number of $n$-cells to a finite cell complex $X^{n-1}$ of dimension $n-1$ using attaching maps $\phi\colon S^{n-1}\to X^{n-1}$.

\noindent The subset $X^k\subset X$ is called the \textbf{k-skeleton}.
\end{definition}

\noindent So a cell complex is formed by inductively gluing n-disks along their boundaries to a cell complex of dimension $n-1$.

\begin{definition}
Let $X^n$ be a finite cell complex. Then $CC_k(X)$ is the free abelian group generated by the $k$-cells $\{e^k_1,...,e^k_{r_k}\}$, so that $CC_k(X)\cong\mathbb{Z}^{r_k}$. $CC_k(X;G)=CC_k(X)\otimes_\mathbb{Z}G$ and $CC^k(X;G)=\text{Hom}(CC_k(X),G)$ for an abelian group $G$.
\end{definition}

\begin{definition}
Let $e_i^k$ be a $k$-cell with an attaching map $\phi_i\colon S^{k-1}\to X^{k-1}$. Given a $(k-1)$-cell $e_j^{k-1}$ we have a map $\phi_i^j:\phi_i^{-1}(e_j^{k-1})\to e_j^{k-1}:x\mapsto\phi_i(x)$. Then We define the boundary map $\partial\colon CC_k(X)\to CC_{k-1(X)}$ by $\partial e_i^k=\sum_jdeg(\phi_i^j)e_j^{k-1}$ and extending linearly.
\end{definition}

\begin{definition}
$d\colon CC^k(X;G)\to CC^{k+1}(X;G)$ is the dual of $\partial$, defined the same way as for singular homology.
\end{definition}

\begin{lemma}
$\partial\circ\partial=0$ and $d\circ d=0$.
\end{lemma}

\begin{definition}
The resultant homology and cohomology groups $H_k(CC_k(X;G),\partial)$ and $H^k(CC^k(X;G),d)$ are called the \textbf{cellular homology} and \textbf{cellular cohomology} groups respectively.
\end{definition}

\begin{proposition}
Let $G$ be an abelian group. Then\\
$H_k(CC_k(X;G),\partial)\cong H_k(X;G)$ and\\
$H^k(CC^k(X;G),d)\cong H^k(X;G)$ for all $k\in\mathbb{N}$.
\end{proposition}

\section{Universal coefficient theorem}
\begin{proposition}
If $G$ is a free abelian group or $G\cong\mathbb{Q}$, then $H_k(X)\otimes_\mathbb{Z}G\cong H_k(X;G)$.\\
If $G=\mathbb{Z}_p$, then $\frac{H_k(X;\mathbb{Z}_p)}{H_k(X)\otimes_\mathbb{Z}\mathbb{Z}_p}\cong T_p(H_{k-1}(X))$.
\end{proposition}

\begin{proposition}
Let $R$ be a commutative ring and let $A$ be an $R$-module.There is a natural inclusion $H_*(X;R)\otimes_RA\xhookrightarrow{}H_*(X;A)$\\
which is an isomorphism if $A$ is free.
\end{proposition}

In particular, $H_*(X;\mathbb{R})\cong H_*(X;\mathbb{Q})\otimes_\mathbb{Q}\mathbb{R}\cong H_*(X)\otimes_\mathbb{Z}\mathbb{R}$. If $H_k(X)$ is finitely generated, then $\text{rk }H_k(X)=\text{dim }H_k(X;\mathbb{R})$

Let $G$ be an abelian group
Recall the natural pairing $H^k(X;G)\times H_k(X)\to G$ given by $([\psi],[c])\mapsto\psi(c)$.

\begin{proposition}
The induced homomorphism $H^k(X;G)\to\text{Hom}(H_k(X);G)$\\
is surjective.
\end{proposition}

Now consider the case $G=\mathbb{Z}$ and set\\
$K=\text{ker}(H^k(X;\mathbb{Z})\to\text{Hom}(H_k(X),\mathbb{Z}))$.
\begin{lemma}
Let $[\psi]\in K$ and $[c]\in H_{k-1}(X)$ such that $n[c]=0$ for some $n\in\mathbb{N}$. Pick $b\in C_k(X)$ such that $nc=\partial b$. Then $\psi(b)\text{ mod }n$ depends only on $[\psi]$ and $[c]$.
In particular, there is a well-defined pairing $K\times T(H_{k-1}(X))\to\mathbb{Q}/\mathbb{Z}$ given by $([\psi],[c])\mapsto\frac{1}{n}\psi(b)$.
\end{lemma}

\begin{proposition}
If $H_*(X)$ is finitely generated, then $K\to\text{Hom}(T(H_{k-1}(X)),\mathbb{Q}/\mathbb{Z})$ is an isomorphism.
\end{proposition}

\begin{remark}
An isomorphism can also be induced by $([\psi],[c])\mapsto\frac{1}{k}\gamma(c)$, where $\gamma\in C^{k-1}(X;\mathbb{Z})$ such that $d\gamma=k\phi$ for some $k\in\mathbb{N}$.
\end{remark}

\begin{corollary}
$K=T(H^k(X;\mathbb{Z}))$.
\end{corollary}
\begin{corollary}
If $H_*(X)$ is finitely generated, then the free part of $H^k(X;\mathbb{Z})$ is naturally dual to the free part of $H_k(X;\mathbb{Z})$ and the torsion part of $H^k(X;\mathbb{Z})$ is naturally dual to the torsion part of $H_{k-1}(X)$.
\end{corollary}

\begin{proposition}
If $A$ is an $R$-module, then the natural homomorphism\\
$H^k(X;A)\to\text{Hom}_R(H_k(X;R),A)$\\
is surjective. If $R$ is a field, then it is an isomorphism.\\
So for a field $\mathbb{F}$, we have $H_k(X;\mathbb{F})\cong H_k(X;\mathbb{F})^*$.
\end{proposition}

\section{Proof of the Borsuk-Ulam theorem with cohomology}

\textrm{(Hatcher section 3.2 exercise 3) \\}
We begin with the following lemmas.

\begin{lemma}
There is no map $\mathbb{R} P^n$$\rightarrow$$\mathbb{R} P^m$ inducing a nontrivial map $H^1(\mathbb{R} P^m;\mathbb{Z}_2)\rightarrow H^1(\mathbb{R} P^n;\mathbb{Z}_2)$ if $n>m$.
\end{lemma}
\begin{proof}
Let $x_n$ and $x_m$ be the generators of $H^1(\mathbb{R} P^n;\mathbb{Z}_2)$ and $H^1(\mathbb{R} P^m;\mathbb{Z}_2)$ respectively.

\[H^1(\mathbb{R} P^n;\mathbb{Z}_2)\cong \mathbb{Z}_2\] 
\[H^*(\mathbb{R} P^n;\mathbb{Z}_2)\cong \mathbb{Z}_2[x]/<x^{n+1}>\] 

\noindent Let $f\colon\mathbb{R} P^n\to\mathbb{R} P^m $ be such a map so that $f^*(x_m)=x_n$.
Then since $f^*$ is a homomorphism on the cup product structure, $f^*(0)=f^*(x_m^{m+1})=f^*(x_m)^{m+1}=x_n^{m+1}=0$, requiring $m \geq n$. The result follows by contraposition.
\end{proof}

\begin{lemma}
If $X$ is a path-connected topological space then $H_1(X;\mathbb{Z}_2)$ is the two-elementary part of $\pi_1(X)^{\text{ab}}$, i.e. the quotient by the subgroup generated by all squares.
\end{lemma}
\begin{proof}
We have that $H_0(X)\cong\mathbb{Z}$ so by the universal coefficient theorem \[\frac{H_1(X;\mathbb{Z}_2)}{H_1(X)\bigotimes_\mathbb{Z}\mathbb{Z}_2}\cong T_2(H_0(X))=0\] The correspondence theorem then gives $H_1(X;\mathbb{Z}_2)=H_1(X)\bigotimes_\mathbb{Z}\mathbb{Z}_2\cong\pi_1(X)^{\text{ab}}\bigotimes_\mathbb{Z}\mathbb{Z}_2$.
\end{proof}

\begin{lemma}
Given any continuous map $f\colon\mathbb{R}P^n\to \mathbb{R}P^m$ with $n>m\geq 0$, $f_*\colon\pi_1(\mathbb{R}P^n)\to \pi_1(\mathbb{R}P^m)$ is trivial.
\end{lemma}
\begin{proof}
 $f^*\colon H^1(\mathbb{R}P^m;\mathbb{Z}_2)\rightarrow H^1(\mathbb{R}P^n;\mathbb{Z}_2)$ is dual to $f_*\colon H_1(\mathbb{R}P^n;\mathbb{Z}_2)\rightarrow H_1(\mathbb{R}P^m;\mathbb{Z}_2)$ which is equal to $f_*\colon \pi_1(\mathbb{R}P^n)\rightarrow \pi_1(\mathbb{R}P^m)$ by the Hurewicz theorem and the fact that $\pi_1(\mathbb{R}P^k)$ is abelian. Thus $f_*$ is trivial.
\end{proof}

\noindent This then entails a proof of the Borsuk-Ulam theorem by applying the ultimate lifting theorem to derive a contradiction.


\section{Lens spaces}
\begin{definition}
The \textbf{3-sphere } $S^3=\{(u,v)\in\mathbb{C}^2:|u|^2+|v|^2=1\}$.
\end{definition}

\begin{definition}
Let $p$ and $q\in\mathbb{N}$ be coprime. Then the \textbf{lens space} $L(p;q)$ is the quotient of $S^3$ by $\mathbb{Z}_p$ with the action generated by $(v,w)\mapsto(e^{2\pi i/p}v,e^{2\pi iq/p}w)$.
\end{definition}

\begin{example}
$L(1;q)\cong S^3$.
\end{example}
\begin{proof}
The group action is generated by $(v,w)\mapsto(e^{2\pi i}v,e^{2\pi iq}w)=(v,w)$ and so is the trivial action. This gives a homeomorphism $S^3\to L(1;q):(v,w)\mapsto[(v,w)]$.
\end{proof}

\begin{example}
$L(2;1)\cong\mathbb{R}P^3$.
\end{example}
\begin{proof}
The group action is generated by $(v,w)\mapsto(e^{\pi i}v,e^{\pi i}w)$ and so $(v,w)\sim(-v,-w)=-(v,w)$ which is the identification which gives $\mathbb{R}P^3$.
\end{proof}

\noindent Thus lens spaces can be considered generalisations of real projective spaces.

\begin{proposition}
$H_1(L(p;q))\cong\pi_1(L(p;q))\cong \mathbb{Z}_p$.
\end{proposition}
\begin{proof}
The multiplication map induced by the group action is clearly continuous for all elements of $\mathbb{Z}_p$.
Now let $(a,b)\in S^3$. If $S=(B_{|a|\text{sin}(\frac{\pi}{100p})}(a)\cap |a|S^1,B_{|b|\text{sin}(\frac{\pi}{100p})}(b)\cap |b|S^1)$ then $gS\cap S=\emptyset$ for all non-trivial $g\in \mathbb{Z}_p$ so the action is a covering space action giving $\pi_1(L(p;q))\cong \mathbb{Z}_p$. $H_1(L(p;q))\cong \mathbb{Z}_p$ by the Hurewicz theorem.
\end{proof}

\noindent Then $H^2(L(p;q);\mathbb{Z})\cong\mathbb{Z}_p$ by Poincare duality. $H_3((L(p;q))\cong\mathbb{Z}\cong H^0(L(p;q);\mathbb{Z})$ since $L(p;q)$ is an orientable 3-manifold. $H_0(L(p;q))\cong\mathbb{Z}\cong H^3(L(p;q);\mathbb{Z})$ since $L(p;q)$ is path-connected. $T(H_2(L(p;q)))\cong T(H^3(L(p;q));\mathbb{Z})=0$ and $F(H_2(L(p;q)))\cong F(H^2(L(p;q);\mathbb{Z}))=0$ by the universal coefficient theorem so $H_2(L(p;q))=0=H^1(L(p;q);\mathbb{Z})$. $H_i(L(p;q))=H^i(L(p;q);\mathbb{Z})=0\forall i>3$ since $L(p;q)$ is a 3-manifold. Thus the homology and cohomology groups depend only on $p$ so are insufficient for a homotopy-equivalence classification.

\begin{definition}
Let $M^n$ be a closed oriented manifold with fundamental class $[M]\in H_n(M;\mathbb{Z})$. Let $[\alpha]\in H^a(M;\mathbb{Z})$ and $[\beta]\in H^{n-a+1}(M;\mathbb{Z})$ be torsion classes, i.e. $k[\alpha]=0,m[\beta]=0$ for some $k,m\in\mathbb{N}$. Then we can write $k\alpha=d\gamma$ for some $\gamma\in C^{a-1}(M,\mathbb{Z})$. We define $[\alpha]\vee[\beta]=\frac{1}{k}(\gamma\cup\beta)M\in\mathbb{Q}/\mathbb{Z}$. The \textbf{torsion linking pairing} is then $T(H^a(M;\mathbb{Z}))\times T(H^{n-a+1}(M;\mathbb{Z}))\to\mathbb{Q}/\mathbb{Z}:([\alpha],[\beta])\mapsto[\alpha]\vee[\beta]$.
\end{definition}
\begin{lemma}
$[\alpha]\vee[\beta]$ is well-defined.
\end{lemma}
\begin{proof}
Let $d\gamma=d\eta=k\alpha$ for $\eta\in C^{a-1}(M,\mathbb{Z})$. Then $d(\gamma-\eta)=0$ so $\gamma-\eta$ is a representative of an element of $H^{a-1}(M;\mathbb{Z})$. $((\gamma-\eta)\cup\beta)M=([\gamma-\eta]\cup[\beta])M$ so $m((\gamma-\eta)\cup\beta)M=([\gamma-\eta]\cup[m\beta])M=0$. Thus $m(\gamma\cup\beta)M=m(\eta\cup\beta)M$ so $(\gamma\cup\beta)M=(\eta\cup\beta)M$, implying independence of $\gamma$.

\noindent Now let $[\alpha]=[\alpha']\in H^a(M;\mathbb{Z})$. Then $\alpha'=\alpha+de$ for some $e\in C^{a-1}(M,\mathbb{Z})$, so $k\alpha'=d\gamma+dke=d(\gamma+ke)$. $\frac{1}{k}((\gamma+ke)\cup\beta)M=\frac{1}{k}(\gamma\cup\beta)M+\frac{1}{k}k(e\cup\beta)M=\frac{1}{k}(\gamma\cup\beta)M \in \mathbb{Q}/\mathbb{Z}$, implying independence of $\alpha$.

\noindent Finally, let $[\beta]=[\beta']\in H^{n-a+1}(M;\mathbb{Z})$. Then $\beta'=\beta+df$ for some $f\in C^{n-a}(M,\mathbb{Z})$. $d(\gamma\cup f)=k(\alpha\cup f)+(-1)^{a-1}(\gamma\cup df)$ so $\frac{1}{k}(\gamma\cup df)M=\frac{(-1)^{1-a}}{k}d(\gamma\cup f)M=\frac{(-1)^{1-a}}{k}([0],[M])=0$. Thus $\frac{1}{k}(\gamma\cup\beta')M=\frac{1}{k}(\gamma\cup(\beta+df))M=\frac{1}{k}(\gamma\cup\beta)M+\frac{1}{k}(\gamma\cup df)M=\frac{1}{k}(\gamma\cup\beta)M$, implying independence of $\beta$. Thus the pairing is well-defined.
\end{proof}

\begin{lemma}
$[\beta]\vee[\alpha]=(-1)^{a(n-a+1)}[\alpha]\vee[\beta]$.
\end{lemma}
\begin{proof}
let $m\beta=d\tau$ for some $\tau\in C^{n-a}(M,\mathbb{Z})$. Then $[\beta]\vee[\alpha]=\frac{1}{m}(\tau\cup\alpha)M=(-1)^{a(n-a)}\frac{1}{m}(\alpha\cup\tau)M=\frac{(-1)^{a(n-a)}}{km}(k\alpha\cup\tau)M=\frac{(-1)^{a(n-a)}}{km}(d\gamma\cup\tau)M$. $d(\gamma\cup\tau)=(d\gamma\cup\tau)+(-1)^{a-1}(\gamma\cup d\tau)=(d\gamma\cup\tau)+(-1)^{a-1}(\gamma\cup m\beta)$ so $\frac{(-1)^{a(n-a)}}{km}(d\gamma\cup\tau)M=\frac{(-1)^{a(n-a)}}{km}(d(\gamma\cup\tau)M-(-1)^{a-1}(\gamma\cup m\beta)M)=\frac{(-1)^{a(n-a)}}{km}(-1)^a(\gamma\cup m\beta)M=\frac{(-1)^{a(n-a+1)}}{k}(\gamma\cup\beta)M=(-1)^{a(n-a+1)}[\alpha]\vee[\beta]$.
\end{proof}

\begin{lemma}
The torsion linking pairing is a non-degenerate bilinear form.
\end{lemma}
\begin{proof}
Bilinearity follows from the bilinearity of the cup product.

\noindent By the universal coefficient theorem, $T(H^{a}(M;\mathbb{Z}))\to\text{Hom}(T(H_{a-1}(M;\mathbb{Z})),\mathbb{Q}/\mathbb{Z}):[\alpha]\mapsto\gamma_{[\alpha]}$ is an isomorphism where $\gamma_{[\alpha]}([l])=\frac{1}{k}\gamma(l)$. By Poincare duality we have an isomorphism $D\colon T(H^{n-a+1}(M;\mathbb{Z}))\to T(H_{a-1}(M;\mathbb{Z})):[\beta]\mapsto[\beta]\cap[M]$ and so $T(H^a(M,\mathbb{Z}))\to \text{Hom}(T(H^{n-a+1}(M;\mathbb{Z})),\mathbb{Q}/\mathbb{Z}):[\alpha]\mapsto\gamma_{[\alpha]}\circ D$ is also an isomorphism where $\gamma_{[\alpha]}\circ D([\beta])=\frac{1}{k}\gamma(\beta\cap M)=\frac{1}{k}(\gamma\cup\beta)M=[\alpha]\vee[\beta]$. Thus $[\alpha]\mapsto ([\beta]\mapsto [\alpha]\vee[\beta])$ is an isomorphism so the bilinear form is non-degenerate.
\end{proof}
\noindent So in the case of 3-manifolds like lens spaces, the torsion linking pairing is a symmetric non-degenerate bilinear form on $T(H^2(M;\mathbb{Z}))$. Also, since the cup product is invariant under homotopy-equivalence, the torsion linking pairing is as well.








\noindent Consider a filled-in ball with points N and S corresponding to the north and south poles and points $x_0,...,x_{p-1}$ being evenly-spaced points along the equator. Then draw a line through the ball from N to S, lines connecting $x_i$ to $x_{i+1}$, lines connecting $x_i$ to N and lines connecting $x_i$ to S (all geodesic). We then identify the faces on the surface by twisting the top hemisphere by $\frac{2\pi q}{p}$ radians and projecting downwards. This leaves us, after identification, with:\\
Points: $[N]$ and $[x_0]$ (since $\{x_0,x_q,x_{2q},...\}$ are identified and $q$ is coprime to $p$)\\
Lines: $[NS]$, $[x_0x_1]$ and $[Nx_0],...,[Nx_{p-1}]$\\
Faces: $[Nx_0x_1],...,[Nx_{p-2}x_{p-1}],[Nx_0x_{p-1}]$ and $[NSx_0],...,[NSx_{p-1}]$\\
Slices: $[NSx_0x_1],...,[NSx_{p-2}x_{p-1}],[NSx_0x_{p-1}]$\\
Let $a_{p-1}=-[NSx_0x_{p-1}]$ and $a_i=[NSx_ix_{i+1}]$ otherwise.\\
Let $b_{p-1}=-[Nx_0x_{p-1}]$ and $b_i=[Nx_ix_{i+1}]$ otherwise.\\
Let $c_i=[NSx_i]$.\\
Let $D=[x_0x_1]$.\\
Let $E=[NS]$.\\
Let $F_i=[Nx_i]$.\\
$-\partial b_{p-1}=\partial[Nx_0x_{p-1}]=[x_0x_{p-1}]-[Nx_{p-1}]+[Nx_0]=[Nx_0]-[Nx_{p-1}]-[x_0x_1]$.\\
Otherwise, $\partial b_i=\partial[Nx_ix_{i+1}]=[x_0x_1]-[Nx_{i+1}]+[Nx_i]$.\\
So $\partial b_i=D-F_{i+1}+F_i$\\
$\partial c_i=\partial[NSx_i]=[Sx_i]-[Nx_i]+[NS]=[NS]-[Nx_i]+[Nx_{i-q}].$\\
So $\partial c_i=E-F_i+F_{i-q}$.\\
$\partial(\sum_{i=0}^{p-1}a_i)=\sum_i([Nx_{i-q}x_{i-q+1}]-[Nx_ix_{i+1}]+[NSx_{i+1}]-[NSx_i])=0$ so $\sum_{i=0}^{p-1}a_i$ is a representative of an element of $H_3(L(p;q);\mathbb{Z})$. In fact, it is a representative of a fundamental class. Call the equivalence class it represents $[M]$.\\
Let $\hat X$ be linear map given by $\hat X(nX)=n$ and $\hat X(Y)=0$ otherwise. We then have\\
$d\hat D=\sum_{i=0}^{p-1}\hat b_i$\\
$d\hat E=\sum_{i=0}^{p-1}\hat c_i$\\
$d\hat F_i=\hat b_i-\hat b_{i-1}-\hat c_i+\hat c_{i+q}$.\\
Then define $\gamma=\hat D + (\sum_{i=0}^{p-1}i\hat F_i)+q\hat E\in C^1(L(p;q);\mathbb{Z})$.\\
$d\gamma=\sum_{i=0}^{p-1}\hat b_i+(\sum_{i=0}^{p-1}i(\hat b_i-\hat b_{i-1}-\hat c_i+\hat c_{i+q}))+\sum_{i=0}^{p-1}q\hat c_i$\\
$=\sum_{i=0}^{p-1}((i+1)\hat b_i-i\hat b_{i-1}-(i-q)\hat c_i+i\hat c_{i+q})\equiv 0\text{ mod }p$.\\
Let $\alpha=\frac{1}{p}d\gamma\in C^2(L(p;q);\mathbb{Z})$. Then $\alpha$ is a representative of an element in $H^2(L(p;q);\mathbb{Z})$ with $p[\alpha]=[0]$.\\
$\gamma(E)=q$.\\
$p\alpha([Sx_{q-1}x_q])=p\alpha([Nx_{p-1}x_0])=p\alpha(b_{p-1})=(p-1+1)\hat b_{p-1}(b_{p-1})-0\times\hat b_{-1}(b_{p-1})=p\implies \alpha([Sx_{q-1}x_q])=1$.\\
Otherwise, $p\alpha([Sx_ix_{i+1}])=p\alpha([Nx_{i-q}x_{i+1-q}])=p\alpha(b_{i-q})=(kp+i-q+1)\hat b_{{i-q}}(b_{{i-q}})-(kp+i-q+1)\hat b_{{i-q}}(b_{{i-q}})=0$, where $k\in\mathbb{Z}$ puts $kp+i-q+1$ in the compatible range for the sum of $[1,p-1]$ (The fact that it can't wrap around to $0$ is why cancellation occurs).\\
Thus $[\alpha]\vee[\alpha]=\frac{1}{p}\gamma([NS])\alpha([Sx_{q-1}x_q])=\frac{q}{p}$.\\

\noindent We then have that if $L(p;q_1)$ and $L(p;q_2)$ are homotopy-equivalent, then $\frac{q_1}{p}=\pm\frac{q_2}{p}m^2$ for some integer $m$, taking into account choice of fundamental class and choice of element of $H^2(L(p;q);\mathbb{Z})$. This is equivalent to $q_1\equiv\pm q_2m^2\text{ mod }p\iff q_1q_2^{-1}\equiv\pm m^2\text{ mod }p\iff q_1q_2\equiv\pm(mq_2)^2\text{ mod } p\iff q_2q_2\equiv\pm n^2\text{ mod } p$, where $n=mq_2$.

If we were only interested in orientation-preserving homotopy-equivalence, then the condition would be $q_2q_2\equiv n^2\text{ mod } p$; and if we were only interested in orientation-reversing homotopy-equivalence, then the condition would be $q_2q_2\equiv- n^2\text{ mod } p$.
Thus $L(p;q)$ is orientation-reversing homotopy-equivalent to itself iff $q(-q)$ is a square mod $p$ $\iff -1$ is a square mod $p\iff p\equiv 1$ mod $4$ for $p$ a prime.

\begin{definition}
Let $X$ be a topological space and $G$ a group acting on it. A \textbf{fundamental domain} is a subset of $X$ which contains precisely one point from each orbit.
\end{definition}

\noindent As such, a fundamental domain is in bijection with $X/G$.

\noindent To construct a lens space as a cell complex we first find a fundamental domain.
$L(p;q)=\{(r_1e^{2\pi i\theta_1},r_2e^{2\pi i\theta_2}):r_1^2+r_2^2=1\}/(\theta_1,\theta_2)\sim(\theta_1+\frac{1}{p},\theta_2+\frac{q}{p})$.

\noindent Let $\rho\colon S^3\to S^3$ represent the action.

\noindent The action rotates the first coordinate by $1/p$ so let $e_k^3=\{(r_1e^{2\pi i\theta_1},r_2e^{2\pi i\theta_2})\in S^3:\frac{k}{p}<\theta_1<\frac{k+1}{p}\}$.

\noindent The $2$-cells will then be the boundaries of the $3$-cells, so let $e_k^2=\{(r_1e^{2\pi i\theta_1},r_2e^{2\pi i\theta_2})\in S^3:\theta_1=\frac{k}{p}\}$.

\noindent Fixing $\theta_1$ leaves $r_2e^{2\pi i\theta_2}$ moving in a circle, so let $e_k^1=\{(r_1e^{2\pi i\theta_1},r_2e^{2\pi i\theta_2})\in S^3:\theta_1=0,\frac{k}{p}<\theta_2<\frac{k+1}{p}\}$.

\noindent Then let $e_k^0=\{(r_1e^{2\pi i\theta_1},r_2e^{2\pi i\theta_2})\in S^3:\theta_1=0,\theta_2=\frac{k}{p}\}$ be the boundary of $e_k^1$.

\noindent Then we have $\rho(e_k^3)=e_{k+1}^3$, $\rho(e_k^2)=e_{k+1}^2$, $\rho(e_k^1)=e_{k+q}^1$ and $\rho(e_k^0)=e_{k+q}^0$

\noindent Thus $\rho$ permutes the cells so taking the quotient gives a cell complex structure $L(p;q)=e^0\cup e^1\cup e^2\cup e^3$.

\noindent $\partial e_0^3=e_1^2-e_0^2=(\rho-\text{id})(e_0^2)$.\\
$\partial e_1^2=e_0^1+...+e_{p-1}^1=(\text{id}+\rho+...+\rho^{p-1})(e_0^1)$.\\
$\partial e_0^1=e_l^0-e_0^0=(\rho^l-\text{id})(e_0^0)$ where $lq\cong1\text{ mod }p$ (so that $l$ corresponds to rotating by $\frac{2\pi i}{p}$).

\noindent Thus on $L(p;q)$, we have $\partial e^3=0$, $\partial e^2=pe^1$, $\partial e^1=0$ since $p=\text{id}$ modulo the group action. This then gives the same homology groups as were found earlier.

\section{Reidemeister torsion}
\begin{definition}
Let $0\to C_n\overset{\partial}{\to}C_{n-1}\overset{\partial}{\to}C_{n-2}\overset{\partial}{\to}...\overset{\partial}{\to}C_{0}\to 0$ be a chain complex with no homology and bases $c_i$ of $C_i$. We can then pick auxiliary bases $b_i\cup\overset{\sim}{b}_{i-1}$ of $C_i$ where $b_i$ is a basis for $\text{Im }\partial_{i+1}$ and $\partial\overset{\sim}{b}_{i-1}=b_{i-1}$. Then let $[b_i\cup\overset{\sim}{b}_{i-1}|c_i]$ be the determinant of the change of basis matrix. We then define The \textbf{Reidemeister torsion} $\Delta(C_.,c_.)=\prod_{i=0}^n[b_i\cup\overset{\sim}{b}_{i-1}|c_i]^{(-1)^i}$.
\end{definition}

\begin{lemma}
$\Delta(C_.,c_.)$ is independent of choice of auxiliary bases.
\end{lemma}
\begin{proof}
Let $b_i'\cup\overset{\sim}{b}_{i-1}'$ be another choice of auxiliary bases. Then 
\end{proof}


\noindent Let $X=\cup_{k=0}^n\cup_je_j^k$ be a finite cell complex and $(A,\beta)$ a representation of $\pi_1(X)$ in a ring $A$. Let $\overset{\sim}{X}$ be a universal cover. And $\overset{\sim}{e}_j^k$ be a lift of $e_j^k$. Then every other lift of $e_j^k$ is obtained by composition of $\overset{\sim}{e}_j^k$ with a deck transformation (since given $f'$ and $g'$ lifts of $f$ there exists a $\phi\in\text{Aut}(p)$ such that $\phi\circ f'(0)=g'(0)$ which implies $\phi\circ f=g$ by uniqueness (lemma 6.9)???). $\text{Aut}(p)=\pi_1(X)$ so we can write $\overset{\sim}{X}=\cup_{k=0}^n\cup_{g\in\pi_1(X)}\cup_jg\overset{\sim}{e}_j^k$. Thus multiplication by an element of $\pi_1(x)$ permutes the cells, thereby giving $C_.^\text{cw}(\overset{\sim}{X})$ a $\pi_1(X)$-action. $A$ also has a $\pi_1(X)$ action so we can define $C_k(X;\beta)=C_k^\text{cw}(\overset{\sim}{X})\otimes_{\mathbb{Z}\pi_1(X)} A$. We have an induced boundary map $\partial\colon C_k(X;\beta)\to C_{k-1}(X;\beta)$ which makes $C_k(X;\beta)$ a chain complex with basis $\{\overset{\sim}{e}_j^k\}\otimes\text{id}_A$.

\begin{definition}
The \textbf{Reidemeister torsion of $X$ with respect to $\beta$} to be $\Delta(X;\beta)=\Delta(C_.(X;\beta)\otimes_{\mathbb{Z}\pi_1(X)}A)$ if $C_.(X;\beta)\otimes_{\mathbb{Z}\pi_1(X)}A$ has no homology. This is a well-defined element of $A^*/\{\pm\beta(\pi_1(X))\}$.
\end{definition}

\noindent This is invariant under homeomorphism but not homotopy-equivalence.



\noindent Let $\omega$ be a $p$th root of unity and let $\beta_\omega\colon\mathbb{Z}_p\to \mathbb{C}^*$ be the representation given by $1\mapsto\omega$.
In $C_.(X;\beta_\omega)\otimes_{\mathbb{Z}\pi_1(X)}\mathbb{C}^*$ we have:\\
$\partial(\overset{\sim}{e}_0^3\otimes\text{id})=(\omega-1)(\overset{\sim}{e}_0^2\otimes\text{id})$\\
$\partial(\overset{\sim}{e}_0^2\otimes\text{id})=(1+\omega+\omega^2+...+\omega^{p-1})(\overset{\sim}{e}_0^1\otimes\text{id})$\\
$\partial(\overset{\sim}{e}_0^1\otimes\text{id})=(\omega^l-1)(\overset{\sim}{e}_0^0\otimes\text{id})$, giving a complex\\
$0\to\mathbb{C}^*\overset{\omega-1}{\to}\mathbb{C}^*\overset{0}{\to}\mathbb{C}^*\overset{\omega^l-1}{\to}\mathbb{C}^*\to0$ whenever $\omega\neq 1$.\\
These maps are isomorphisms and so the complex has no homology. Hence for $\omega\neq 1$ we have $\Delta(L(p;q);\beta_\omega)=(\omega^l-1)(\omega-1)\in\mathbb{C}^*/\pm\omega^k$ for $k\in\mathbb{Z}$.

\noindent $(\omega^l-1)(\omega-1)=\omega^l(1-\omega^{-l})(\omega-1)=(\omega^{-l}-1)(\omega-1)=\omega(\omega^{-l}-1)(1-\omega^{-1})=(\omega^{-l}-1)(\omega^{-1}-1)$ So $\Delta(L(p;q);\beta_\omega)=\Delta(L(p;-q);\beta_\omega)$.\\
$(\omega^l-1)(\omega-1)=(\omega^q-1)(\omega-1)$ so $\Delta(L(p;q);\beta_\omega)=\Delta(L(p;q^{-1});\beta_\omega)$.\\
Thus two lens spaces $L(p;q_1)$ and $L(p;q_2)$ can only be homeomorphic if $q_1\equiv \pm q_2^{\pm 1}\text{mod }p$.

\begin{theorem}
Let $L(p;q_1)$ and $L(p;q_2)$ be lens spaces.\\
They are homotopy equivalent if and only if $q_1q_2\equiv\pm n^2 \text{ mod } p$ for some $n\in\mathbb{N}$.\\
They are homeomorphic if and only if $q_1\equiv\pm q_2^{\pm 1} \text{ mod }p$.
\end{theorem}

\begin{example}
$L(7;1)$ and $L(7;2)$ are homotopy-equivalent but not homeomorphic.
\end{example}
\end{document}

